---
topic: "air"
title: Status of Ground-Level Ozone in B.C. (2014-2016)
output: envreportbc::print_ver
---

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, 
                      error=FALSE)

library("dplyr") # data munging
library("ggplot2") # for plotting
library("scales") # for date_breaks()
library("envreportutils") # for theme_facet_soe()
library("forcats") # tweak factor levels fct_drop()
library("sp") # mapping
library("knitr") #for knitting to PDF

## Reading in datasets
load("../tmp/analysed.RData")

read_chunk("../04_output.R")

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # back slash has to be first
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}

```

```{r pre}
```

British Columbia operates a network of air monitoring stations that
measure ground-level ozone and other air pollutants. This indicator
reports on the concentration of ground-level ozone from 2014-2016 and
compares it to the [Canadian Ambient Air Quality Standard](http://www.ccme.ca/en/resources/air/aqms.html#anchor1)
established by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

-   **Ground-level ozone is an air pollutant.** While ozone in the
    Earth's atmosphere occurs naturally, additional ozone at the
    ground level is a pollutant. It forms through chemical reactions
    involving nitrogen oxides and volatile organic compounds in the
    presence of sunlight.
-   **Ground-level ozone can be harmful to humans.** Exposure to
    ground-level ozone (hereafter ozone) can reduce lung function and
    cause inflammation of airways, which can increase respiratory
    symptoms and aggravate asthma^[1](#fn1)^. These effects are linked
    to more emergency room visits, hospitalizations, and absenteeism, and
    higher health care costs^[2](#fn2)^.
-   **Ozone levels met the Canadian Ambient Air Quality Standard at all B.C. 
    ozone monitoring stations.** The ozone metric (see sidebar) from the 39 
    stations ranged from 34 to 60 parts per billion (ppb).
    Ozone metric values were $\leq\!\!$ 50 ppb at 26 reporting stations (67%). No
    stations reported CAAQs metric values \>60 ppb.
-   **Ozone levels met the Canadian Ambient Air Quality Standard in six
    of B.C.'s seven air zones.** The ozone level for an air zone is the 
    highest ozone metric value reported from monitoring stations within 
    the air zone. Currently, there are no [air monitoring stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest Air Zone.
-   **Ozone metrics are used to set management levels for each air
    zone.** Four management levels (green, yellow, orange, and red) are each
    associated with a suite of actions that become more rigorous as
    ozone metric values approach the Canadian Ambient Air Quality
    Standard.


\begin{center}
\fbox{
  \begin{minipage}[c]{15cm}
  
    \begingroup\centering{\textbf{Ground-Level Ozone: Canadian Ambient Air Quality Standard}}
    
    \endgroup
    
    \vspace{1mm}
    
    \text{The Canadian Ambient Air Quality Standard ozone metric should be \textbf{lower than 63 parts per billion}.}
    
    \href{http://www.ccme.ca/en/resources/air/aqms.html}{Learn more about the  Canadian Ambient Air Quality Standards.}
    
  \end{minipage}
}
\end{center}



\newpage

The map and chart below summarise Canadian Ambient Air Quality Standard
achievement status for ozone in B.C. air zones, as well as the ozone metric at individual monitoring stations. Summaries are given for each monitoring station where sufficient data was available for the 2014-2016
reporting period.

# Status of Ozone Levels in B.C. Air Zones (2014-2016)

```{r achievement_map, fig.height=8, fig.width=7}
```

\newpage

# Ozone Metric for Air Monitoring Stations within B.C. Air Zones (2014-2016)

```{r summary_plot, fig.height=9, fig.width=7}
```



### More about the Canadian Ambient Air Quality Standard ozone metric:

-   The Canadian Ambient Air Quality Standard ozone metric is the 3-year average of the annual 4th-highest daily          maximum 8-hour rolling average concentration.
-   The ozone metric for an air zone is the highest ozone metric value reported from monitoring stations within the       air zone. Air zone metric values are only taken from stations reporting on two years of data when there      are no available three year reporting stations in the air zone.
-   Only air monitoring stations with sufficient data on ozone
    concentrations for the 2014-2016 reporting period were included in
    this indicator. Visit [Current Air Quality Data](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E) for a
    complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the ozone
    metric are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).

## Canada-wide Air Quality Management System

The Air Quality Management System is Canada's approach for protecting air quality. Under the Air Quality Management System, provinces and territories monitor, report and manage local air quality within air zones, with the goal of continuously improving air quality and keeping pollutants below the Canadian Ambient Air Quality Standards.

-   Provinces and territories establish air zones that are defined by a similar set of air quality characteristics,       issues and trends. There are seven air zones in British Columbia: Coastal, Georgia Strait, Lower Fraser Valley,       Southern Interior, Central Interior, Northeast and Northwest.
-   The Air Zone Management Framework defines management levels based on the highest ozone metric value reported from     monitoring stations within the air zone. The four management levels are (1) green (actions for keeping clean areas     clean); (2) yellow (actions for preventing air quality deterioration); (3) orange (actions for preventing Canadian     Ambient Air Quality Standard exceedance); and (4) red (actions for achieving air zone Canadian Ambient Air Quality     Standard).
-   All six B.C. air zones that were assessed met the Canadian Ambient Air
    Quality Standard for ground-level ozone ($\leq\!\!$ 63 ppb).
-   Based on the highest ozone metric value reported within each air zone, the Coastal air zone was assigned to a         green management level, the Northeast, Central Interior, Southern Interior and Georgia Strait air zones were          assigned to a yellow management level, and the Lower Fraser Valley was assigned to an orange management level.        Currently, there are no [air monitoring     stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E) in the Northwest air zone.
-   Each management level is associated with a suite of actions that become more rigorous as ozone             concentration levels approach the Canadian Ambient Air Quality Standard. Actions for each management level are detailed in the [Guidance Document on Air Zone Management
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf).

# Management Levels for Ozone in B.C. Air Zones (2014-2016)

```{r mgmt_map, fig.height = 5}
```


```{r mgmt_chart}
```
```{r fig.height=4}
mgmt_chart +  theme(plot.margin = unit(c(10,30,0,0),"mm"))
```


### More about the Air Quality Management System management levels:

-   There can be cases when an air zone does not achieve a given Canadian Ambient Air Quality Standard
    because of sources over which jurisdictions have little or no
    control, such as those related to transboundary flows and
    exceptional events, like forest fires.
-   Prior to determining management levels, jurisdictions have the
    option of adjusting their air zone metric values to remove such
    external influences. These arrangements aim to ensure that
    jurisdictions are responsible for managing only the emissions
    sources they can control.
-   Two days were flagged as exceptional events as a result of suspected wildfire influence in B.C. for the 2014-2016 reporting period (July 8-9, 2015 for [Agassiz Municipal Hall](http://www.env.gov.bc.ca/epd/bcairquality/readings/map/station.html#E293810)). However, removal of the data from these days had no impact on the resulting air zone management levels for ground-level ozone.

# Methods

The methods used to develop this indicator --- including procedures, data
requirements, and calculation of the Canadian Ambient Air Quality Standard ozone metric --- are detailed in
the [Guidance Document on
Achievement Determination: Canadian Ambient Air Quality Standards for
Fine Particulate Matter and Ozone
(2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf) published by the [Canadian Council of Ministers of the Environment](http://www.ccme.ca/index.html).

**R package and code:** We have developed an [R](http://www.r-project.org) package to facilitate the calculation of air quality metrics according to the [Canadian Ambient Air Quality Standards](http://www.ccme.ca/en/current_priorities/air/caaqs.html). Download the 'rcaaqs' package from [GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating the analysis presented on this page is also available on [GitHub](https://github.com/bcgov/ozone_caaqs_analysis).


# References and Other Useful Links

[Learn more about the implementation of the Air Quality Management System in British Columbia](http://www2.gov.bc.ca/gov/content?id=8FA84ADC845A49AB93E525A58D9D4A97)

Read individual [Air Zone reports](http://www2.gov.bc.ca/gov/search?q=air+zone%2Binmeta%3AnrsAirQuality_nrs_air_quality_topic%3DAir+Zone+Reports&id=04D14E7F7EE043889E1AC20E205CA3F8&tab=1&sourceId=267EE7E2432B4DE89BF1905EF80F505E) on the achievement of the Canadian Ambient Air Quality Standards for ground-level ozone and fine particulate matter in B.C.

[Access B.C.'s real-time data on air pollutants and find locations of all provincial air monitoring stations in B.C.](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)

[BC Lung Asssociation's BC State of the Air Reports](http://www.bc.lung.ca/airquality/stateoftheair-report.html)

[Canadian Environmental Sustainability Indicators: Air and Climate Indicators](http://www.ec.gc.ca/indicateurs-indicators/default.asp?lang=En&n=03603FB3-1)

^1^[United States Environmental Protection Agency. February 2013.
Integrated Science Assessment of Ozone and Related Photochemical
Oxidants
(600/R-10/076F)](http://cfpub.epa.gov/ncea/isa/recordisplay.cfm?deid=247492#Download)

^2^Willey, J., N. Gilbert, and N. Lyrette. 2004. Human health effects of
ozone: Update in support of the Canada-wide standards for particulate
matter and ozone. Revised version. Working paper prepared for Canadian
Council of Ministers of the Environment. Health Canada. Ottawa, Ontario:
Health Canada

# Data

\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Indicator data: BC Ground-Level Ozone Canadian Ambient Air Quality Standards 2014-2016](https://catalogue.data.gov.bc.ca/dataset/9b7a9e74-9274-4f97-be81-ce4ee475077d) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [BC Ozone Hourly Data & Air Monitoring Station Locations](https://catalogue.data.gov.bc.ca/dataset/77eeadf4-0c19-48bf-a47a-fa9eef01f409) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [BC Air Zones](https://catalogue.data.gov.bc.ca/dataset/e8eeefc4-2826-47bc-8430-85703d328516) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (August 2017):
<http://www.env.gov.bc.ca/soe/indicators/air/ozone.html>

Email correspondence to: envreportbc@gov.bc.ca

\newpage

# Appendix: Detailed ground-level ozone Canadian Ambient Air Quality Standard results for each monitoring station within air zones in B.C.


```{r appendix, results='asis', fig.height=3.5}

mid_breaks <- function(width = "1 year") {
  function(x) {
    if (length(x) > 2) stop("x should be a range of length 2")
    sq <- scales::fullseq(x, width)
    diff <- diff(sq)
    sq[-length(sq)] + diff / 2
  }
}


ems_ids <- sum_dat$ems_id
sum_dat$caaq_status <- as.character(sum_dat$caaq_status)

plot_exceedances <- FALSE

for (az in (unique(sum_dat$Airzone))) {
   dat <- sum_dat[sum_dat$Airzone == az, ]
   ems_ids <- unique(dat$ems_id)
   cat("\\subsection{", az, " Air Zone}", sep = "")

for (emsid in ems_ids) {
  site_sum <- dat[dat$ems_id == emsid, ]
  dailydata <- daily_max_o3[daily_max_o3$ems_id == emsid &
                              daily_max_o3$date <= maxdate,]
  
  if (nrow(dailydata) == 0) next
  
 # site <- dailydata$station_name[1]
  
  site <- knitr_latex_char(site_sum$station_name[1])
  cat("\\subsubsection*{\\underline{", site, " monitoring station}}", sep = "")
  cat("\\textbf{Ozone Air Quality Standard:} ", site_sum$caaq_status, " \\newline ")
  cat("\\textbf{Ozone Metric:} ", site_sum$caaq_metric, " ppb (", site_sum$caaq_nYears, " year average) \\newline ", sep = "")
  
 caaq_data <- sum_dat[sum_dat$ems_id == emsid,]

  
  lineplot <- ggplot(dailydata, size = 1) + 
    scale_x_date(expand = c(0,50), limits = c(mindate - 1, maxdate), 
                 breaks = mid_breaks(), labels = date_format("%Y")) + 
    scale_y_continuous(limits = c(0, 102), 
                       breaks = seq(0, 100, by = 20), labels = seq(0, 100, by = 20), 
                       expand = c(0,0)) + 
    geom_line(aes(x = date, y = max8hr), colour = "#9ecae1", size = 0.5) + 
    geom_hline(aes(yintercept = o3_standard), linetype = 2, colour = "#e41a1c") + 
    annotate("text", label = paste0("Ozone Standard (", o3_standard, " ppb)  "), 
             x = maxdate, y = o3_standard + 3, vjust = 0, hjust = 1, 
             size = 3.5, colour = "#e41a1c") + 
    theme_soe(base_size = 10) + 
    theme(axis.title.y = element_text(vjust = 1),
          plot.margin = unit(c(10,0,0,0),"mm"),
              panel.grid.major.x = element_blank(), 
          panel.grid.minor.x = element_line(colour = "grey85")) + 
    labs(x = NULL, y = "Daily Maximum Ozone\n(parts per billion)")
  
  if (plot_exceedances) {
    exceedance_data <- filter(dailydata, exceed)
    
    if (nrow(exceedance_data) > 0) {
      lineplot <- lineplot + 
        geom_point(data = exceedance_data, aes(x = date, y = max8hr), 
                   colour = "#e41a1c", size = 2) + 
        annotate("text", x = exceedance_data$date[1] + 20, y = exceedance_data$max8hr[1], 
                 label = "Exceedances", hjust = 0, vjust = 0, colour = "#e41a1c", size = 3)
    }
  }
    
  if (nrow(caaq_data) > 0) {
    lineplot <- lineplot + 
      geom_segment(data = caaq_data, 
                   mapping = aes(x = as.Date(paste0(caaq_year_min, "-01-01")), 
                                 xend = as.Date(paste0(caaq_year_max,"-12-31")), 
                                 y = caaq_metric, yend = caaq_metric, 
                                 colour = factor(caaq_status, 
                                                 levels = c("Achieved", "Not Achieved"))), 
                   size = 1.5) + 
      annotate("text", x = as.Date(paste0(caaq_data$caaq_year_min, "-01-30")), 
               y = 73, label = "2014-2016 Ozone Metric", size = 3.5, hjust=0, 
               colour = "grey50") + 
      geom_segment(data = caaq_data, colour = "grey60",
                   aes(x = as.Date(paste0(caaq_year_min,"-09-15")), y = 69, 
                       xend = as.Date(paste0(caaq_year_min, "-11-01")), 
                       yend = caaq_metric + 1)) +
      scale_colour_manual(values = c("#377eb8", "#e41a1c"), labels = "2014-2016 Ozone Metric", 
                          name = element_blank(), guide = "none") 

  }
  
  
  plot(lineplot)
  cat("\\newpage ")
  }
   
}


```


