---
topic: "air"
title: Status of Ground-Level Ozone in B.C. (2014-2016)
output: envreportbc::print_ver
---

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, 
                      error=FALSE)

library("dplyr") # data munging
library("ggplot2") # for plotting
library("scales") # for date_breaks()
library("envreportutils") # for theme_facet_soe()
library("forcats") # tweak factor levels fct_drop()
library("sp") # mapping
library("knitr") #for knitting to PDF

## Reading in datasets
load("../tmp/analysed.RData")

read_chunk("../04_output.R")

```

```{r pre}
```

British Columbia operates a network of air monitoring stations that
measure ground-level ozone and other air pollutants. This indicator
reports on the concentration of ground-level ozone from 2014-2016 and
compares it to the [Canadian Ambient Air Quality Standard](http://www.ccme.ca/en/resources/air/aqms.html#anchor1)
established by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

-   **Ground-level ozone is an air pollutant.** While ozone in the
    Earth's atmosphere occurs naturally, additional ozone at the
    ground level is a pollutant. It forms through chemical reactions
    involving nitrogen oxides and volatile organic compounds in the
    presence of sunlight.
-   **Ground-level ozone can be harmful to humans.** Exposure to
    ground-level ozone (hereafter ozone) can reduce lung function and
    cause inflammation of airways, which can increase respiratory
    symptoms and aggravate asthma^[1](#fn1)^. These effects are linked
    to more emergency room visits, hospitalizations, and absenteeism, and
    higher health care costs^[2](#fn2)^.
-   **Ozone levels met the Canadian Ambient Air Quality Standard at all B.C. 
    ozone monitoring stations.** The ozone metric (see sidebar) from the 39 
    stations ranged from 34 to 60 parts per billion (ppb).
    Ozone metric values were $\leq\!\!$ 50 ppb at 26 reporting stations (67%). No
    stations reported CAAQs metric values \>60 ppb.
-   **Ozone levels met the Canadian Ambient Air Quality Standard in six
    of B.C.'s seven air zones.** The ozone level for an air zone is the 
    highest ozone metric value reported from monitoring stations within 
    the air zone. Currently, there are no [air monitoring stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest Air Zone.
-   **Ozone metrics are used to set management levels for each air
    zone.** Four management levels (green, yellow, orange, and red) are each
    associated with a suite of actions that become more rigorous as
    ozone metric values approach the Canadian Ambient Air Quality
    Standard.


\begin{center}
\fbox{
  \begin{minipage}[c]{15cm}
  
    \begingroup\centering{\textbf{Ground-Level Ozone: Canadian Ambient Air Quality Standard}}
    
    \endgroup
    
    \vspace{1mm}
    
    \text{The Canadian Ambient Air Quality Standard ozone metric should be \textbf{lower than 63 parts per billion}.}
    
    \href{http://www.ccme.ca/en/resources/air/aqms.html}{Learn more about the  Canadian Ambient Air Quality Standards.}
    
  \end{minipage}
}
\end{center}



\newpage

The map and chart below summarise Canadian Ambient Air Quality Standard
achievement status for ozone in B.C. air zones, as well as the ozone metric at individual monitoring stations. Summaries are given for each monitoring station where sufficient data was available for the 2014-2016
reporting period.

```{r achievement_map}
```
```{r fig.height=8, fig.width=7}
caaqs_achievement_map <- caaqs_achievement_map + labs(title = "Status of Ozone Levels in B.C. Air Zones (2014-2016)")
plot(caaqs_achievement_map)
```

\newpage

```{r summary_plot}
```
```{r fig.height=9, fig.width=7}
summary_plot <- summary_plot + labs(title = "Ozone Metric for Air Monitoring Stations within B.C. Air Zones")
plot(summary_plot)
```


### More about the Canadian Ambient Air Quality Standard ozone metric:

-   The Canadian Ambient Air Quality Standard ozone metric is the 3-year average of the annual 4th-highest daily maximum 8-hour rolling average concentration.
-   The ozone metric for an air zone is the highest ozone metric value reported from monitoring stations within the air zone. However, air zone metric values are only taken from stations reporting on two years of data when there are no available three year reporting stations in the air zone.
-   Only air monitoring stations with sufficient data on ozone
    concentrations for the 2014-2016 reporting period were included in
    this indicator. Visit [Current Air Quality Data](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E) for a
    complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the ozone
    metric are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).

## Canada-wide Air Quality Management System

The Air Quality Management System is Canada's approach for protecting air quality. Under the Air Quality Management System, provinces and territories monitor, report and manage local air quality within air zones, with the goal of continuously improving air quality and keeping pollutants below the Canadian Ambient Air Quality Standards.

-   Provinces and territories establish air zones that are defined by a
    similar set of air quality characteristics, issues and trends.
-   There are seven air zones in British Columbia: Coastal, Georgia Strait,
    Lower Fraser Valley, Southern Interior, Central Interior, Northeast
    and Northwest.
-   The Air Zone Management Framework defines management levels based on
    the highest CAAQS ozone metric value reported from monitoring stations
    within the air zone.
-   The four management levels include: (1) green (actions for keeping
    clean areas clean); (2) yellow (actions for preventing air quality
    deterioration); (3) orange (actions for preventing CAAQS
    exceedance); and (4) red (actions for achieving air zone CAAQS). Each management level is     associated with a suite of actions that become more rigorous as ozone concentration levels     approach
    the CAAQS.
-   All four B.C. air zones that were assessed met the Canadian Ambient Air
    Quality Standard for ground-level ozone ($\leq\!\!$ 63 ppb).
-   Based on the highest CAAQS ozone metric value reported within each air zone,
    the Central Interior, Georgia Strait, and Lower Fraser Valley air zones
    were assigned to a yellow management level (actions for preventing air
    quality deterioration) and the Southern Interior air zone to an orange
    management level (actions for preventing CAAQS exceedance). Due to
    insufficient data, ozone management levels in the Coastal, Northwest and
    Northeast air zones are yet to be determined.
-   Actions for each management level are detailed in the
    [CCME](http://www.ccme.ca/en/current_priorities/air/caaqs.html)
    [Guidance Document on Air Zone Management
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf).

<!-- ```{r mgmt_map, fig.height = 5} -->
<!-- colrs <- c("Insufficient Data" = "grey80",  -->
<!--            "Actions for Keeping Clean Areas Clean" = "#A6D96A",  -->
<!--            "Actions for Preventing Air Quality Deterioration" = "#FEE08B",  -->
<!--            "Actions for Preventing CAAQS Exceedance" = "#F46D43",  -->
<!--            "Actions for Achieving Air Zone CAAQS" = "#A50026") -->

<!-- mgmt_map <- ggplot(airzones.points, aes(long, lat)) +    -->
<!--   geom_polygon(aes(group=group, fill = caaq_level)) +  -->
<!--   coord_fixed() +  -->
<!--   geom_path(aes(group=group), colour = "white") +  -->
<!--   ggtitle("Management Levels for Ozone in B.C. Air Zones") +  -->
<!--   theme_minimal() +  -->
<!--   scale_fill_manual(values = colrs,  -->
<!--                     drop = FALSE,  -->
<!--                     name = "Air Zone Management Levels",  -->
<!--                     guide = guide_legend(reverse = TRUE)) +  -->
<!--   theme(axis.title = element_blank(), -->
<!--         axis.text = element_blank(),  -->
<!--         axis.ticks = element_blank(), -->
<!--         panel.grid = element_blank(), -->
<!--        legend.position = "none", -->
<!--        plot.margin = unit(c(0,0,0,0),"mm")) + -->
<!--   annotate("text", x=680000, y=950000, label="Coastal",  -->
<!--            colour="black", size=3.5) + -->
<!--   annotate("text", x=1150000, y=1550000, label="Northeast",  -->
<!--            colour="black", size=3.5) + -->
<!--   annotate("text", x=780000, y=1500000, label="Northwest",  -->
<!--            colour="black", size=3.5) + -->
<!--   annotate("text", x=1150000, y=950000, label="Central\n Interior",  -->
<!--            colour="black", size=3.5) + -->
<!--   annotate("text", x=1550000, y=600000, label="Southern\n Interior",  -->
<!--            colour="black", size=3.5) + -->
<!--   annotate("text", x=1150000, y=325000, label="Georgia Strait",  -->
<!--            colour="black", size=3.5) + -->
<!--   annotate("text", x=1500000, y=410000, label="Lower Fraser Valley",  -->
<!--            colour="black", size=3.5) -->

<!-- plot(mgmt_map) -->
<!-- ``` -->

<!-- ```{r mgmt_chart, fig.height=4} -->
<!-- chart_colrs <- colrs[-1] -->

<!-- # <- c(expression(paste("Actions for Keeping Clean Areas Clean ("<=50, " ppb)")),  -->
<!-- #             expression(paste("Actions for Preventing Air Quality Deterioration (">50, " & " <= 56, " #ppb)")),  -->
<!-- #             expression(paste("Actions for Preventing CAAQS Exceedance (">56, " & " <=63, " ppb)")),  -->
<!-- #             expression(paste("Actions for Achieving Air Zone CAAQS (">63, " ppb)"))) -->

<!-- mlevels <- c("Actions for Keeping Clean Areas Clean (<=50 ppb)",  -->
<!--              "Actions for Preventing Air Quality Deterioration (>50 & <=56 ppb)",  -->
<!--              "Actions for Preventing CAAQS Exceedance (>56 & <=63 ppb)",  -->
<!--              "Actions for Achieving Air Zone CAAQS (>63 ppb)") -->

<!-- mgmt_chart <- ggplot(data=stations.points, -->
<!--                               aes(x = Airzone, fill = caaq_level)) +  -->
<!--   geom_bar(stat = "bin", alpha = 1) + -->
<!--   xlab ("") + ylab ("Number of Reporting Stations") + -->
<!--   scale_y_continuous(limits = c(0,20), breaks=seq(0, 20, 4), -->
<!--                      expand=c(0,0)) + -->
<!--   scale_fill_manual(values = chart_colrs,  -->
<!--                     drop = FALSE,  -->
<!--                     name = "Air Zone Management Levels",  -->
<!--                     guide = guide_legend(reverse = TRUE), -->
<!--                     labels = mlevels) +  -->
<!--   theme_soe(base_family = "") + -->
<!--   theme(panel.grid.major.y = (element_blank()), -->
<!--         axis.text = element_text(size = 12), -->
<!--         legend.position = "bottom", -->
<!--         legend.direction = "vertical", -->
<!--         legend.margin = unit(15,"mm"), -->
<!--         legend.text.align = 0) + -->
<!--   coord_flip() -->

<!-- plot(mgmt_chart) -->
<!-- ``` -->


<!-- ### More about the AQMS management levels: -->

<!-- -   There can be cases when an air zone does not achieve a given CAAQS -->
<!--     because of sources over which jurisdictions have little or no -->
<!--     control, such as those related to transboundary flows and -->
<!--     exceptional events like forest fires. -->
<!-- -   Prior to determining management levels, jurisdictions have the -->
<!--     option of adjusting their air zone metric values to remove such -->
<!--     external influences. These arrangements aim to ensure that -->
<!--     jurisdictions are responsible for managing only the emissions -->
<!--     sources they can control. -->
<!-- -   For the 2011-2013 reporting period, all ozone monitoring stations met -->
<!--     the CAAQS and no adjustments for transboundary or exceptional events were -->
<!--     warranted. Air zone management levels for ground-level ozone were -->
<!--     determined directly from the CAAQS ground-level ozone metric for this -->
<!--     reporting period.  -->

<!-- # Methods -->

<!-- The methods used to develop this indicator --- including procedures, data -->
<!-- requirements, and calculation of the CAAQS ozone metric --- are detailed in -->
<!-- the [CCME](http://www.ccme.ca/index.html) [Guidance Document on -->
<!-- Achievement Determination: Canadian Ambient Air Quality Standards for -->
<!-- Fine Particulate Matter and Ozone -->
<!-- (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).  -->
<!-- **R package and code:** We have developed an [R](http://www.r-project.org) package to facilitate the calculation of air quality metrics according to the Canadian Ambient Air Quality Standards [CAAQS](http://www.ccme.ca/en/current_priorities/air/caaqs.html). Download the 'rcaaqs' package from [GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating the analysis presented on this page is also available on [GitHub](https://github.com/bcgov/ozone_caaqs_analysis). -->


<!-- # References and Other Useful Links -->

<!-- Visit [BC Air Quality](http://www.bcairquality.ca/plans/national-air-quality-management-system.html) for more information on the implementation of the AQMS in B.C., and to read individual Air Zone reports on the achievement of CAAQS for ground-level ozone and fine particulate matter. -->

<!-- Access [B.C.'s Air Quality -->
<!-- Readings](http://www.bcairquality.ca/readings/index.html) for real-time -->
<!-- data on air pollutants and locations of all air monitoring stations in -->
<!-- B.C. -->

<!-- For more details on the CAAQS and Canada's Air Quality Management System -->
<!-- visit [Canada-wide Air Quality Management System -->
<!-- (AQMS)](http://www.ccme.ca/en/resources/air/aqms.html) -->

<!-- [BC Lung Asssociation's BC State of the Air Reports](http://www.bc.lung.ca/airquality/stateoftheair-report.html) -->

<!-- ^1^[United States Environmental Protection Agency. February 2013. -->
<!-- Integrated Science Assessment of Ozone and Related Photochemical -->
<!-- Oxidants -->
<!-- (600/R-10/076F)](http://cfpub.epa.gov/ncea/isa/recordisplay.cfm?deid=247492#Download) -->

<!-- ^2^Willey, J., N. Gilbert, and N. Lyrette. 2004. Human health effects of -->
<!-- ozone: Update in support of the Canada-wide standards for particulate -->
<!-- matter and ozone. Revised version. Working paper prepared for Canadian -->
<!-- Council of Ministers of the Environment. Health Canada. Ottawa, Ontario: -->
<!-- Health Canada -->

<!-- # Data -->

<!-- \*By accessing these datasets, you agree to the license associated with -->
<!-- each file, as indicated in parentheses below. -->

<!-- -   [Indicator data: BC Ozone CAAQS 2011-2013](http://catalogue.data.gov.bc.ca/dataset/bc-ambient-air-quality-standards-reporting-results-ozone) ([License: B.C. -->
<!--     OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf)) -->
<!-- -   [Hourly ozone data](http://catalogue.data.gov.bc.ca/dataset/air-quality-monitoring-verified-hourly-data-and-station-data) (License: [B.C.     OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf)) -->
<!-- -   [BC Air Monitoring     Stations](http://catalogue.data.gov.bc.ca/dataset/air-quality-monitoring-verified-hourly-data-and-station-data) -->
<!--     (License: [B.C. -->
<!--     OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf)) -->
<!-- -   [BC Air Zones](http://catalogue.data.gov.bc.ca/dataset/british-columbia-air-zones) (License: [B.C. -->
<!--     OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf)) -->

<!-- Published and Available On-Line at Environmental Reporting BC (August 2015): -->
<!-- <http://www.env.gov.bc.ca/soe/indicators/air/ozone.html> -->

<!-- Email correspondence to: envreportbc@gov.bc.ca -->

<!-- \newpage -->

<!-- # Appendix: Detailed CAAQS results for each monitoring station within air zones in B.C. -->

<!-- ```{r appendix, results='asis', fig.height=3.5} -->
<!-- plot_exceedances <- FALSE -->

<!-- for (az in (unique(sum_dat$Airzone))) { -->
<!--   dat <- sum_dat[sum_dat$Airzone == az, ] -->
<!--   ems_ids <- unique(dat$ems_id) -->
<!--   cat("\\subsection{", az, " Air Zone}", sep = "") -->

<!--   for (emsid in ems_ids) { -->
<!--     site_sum <- dat[dat$ems_id == emsid, ] -->
<!--     dailydata <- daily_8hr_roll_max[daily_8hr_roll_max$ems_id == emsid & -->
<!--                                       daily_8hr_roll_max$date <= maxdate,] -->

<!--     if (nrow(dailydata) == 0) next -->

<!--     site <- knitr_latex_char(site_sum$display_name[1]) -->
<!--     cat("\\subsubsection*{\\underline{", site, " monitoring station}}", sep = "") -->
<!--     cat("\\textbf{Ozone Air Quality Standard:} ", site_sum$caaq_status, " \\newline ") -->
<!--     cat("\\textbf{Ozone Metric:} ", site_sum$caaq_metric, " ppb (", site_sum$caaq_nYears,  -->
<!--         " year average) \\newline ", sep = "") -->
<!--     cat("\\textbf{Air Quality Management Level:} ", as.character(site_sum$caaq_level),  -->
<!--         " \\newline \\newline \\newline ") -->

<!--     caaq_data <- three_yr_avg[three_yr_avg$ems_id == emsid,] -->

<!--   lineplot <- ggplot(dailydata, size = 1) +  -->
<!--     scale_x_date(expand = c(0,50), limits = c(mindate - 1, maxdate),  -->
<!--                  breaks = date_breaks(width = "1 year"), labels = date_format("%Y")) +  -->
<!--   scale_y_continuous(limits = c(0, 102),  -->
<!--                      breaks = seq(0, 100, by = 20), labels = seq(0, 100, by = 20),  -->
<!--                      expand = c(0,0)) +  -->
<!--   geom_line(aes(x = date, y = max8hr), colour = "#9ecae1", size = 0.5) +  -->
<!--     geom_hline(aes(yintercept = o3_standard), linetype = 2, colour = "#e41a1c") +  -->
<!--     annotate("text", label = paste0("Ozone Standard (", o3_standard, " ppb)  "),  -->
<!--              x = maxdate, y = o3_standard + 3, vjust = 0, hjust = 1,  -->
<!--              size = 3.5, colour = "#e41a1c") +  -->
<!--   theme_soe(base_family = "", base_size = 10) +  -->
<!--     theme(axis.title.y = element_text(vjust = 1)) +  -->
<!--     labs(x = NULL, y = "Daily Maximum Ozone\n(parts per billion)") -->

<!--   if (plot_exceedances) { -->
<!--     exceedance_data <- filter(dailydata, exceed) -->

<!--     if (nrow(exceedance_data) > 0) { -->
<!--       lineplot <- lineplot +  -->
<!--         geom_point(data = exceedance_data, aes(x = date, y = max8hr),  -->
<!--                    colour = "#e41a1c", size = 2) +  -->
<!--         annotate("text", x = exceedance_data$date[1] + 20, y = exceedance_data$max8hr[1],  -->
<!--                  label = "Exceedances", hjust = 0, vjust = 0, colour = "#e41a1c", size = 3) -->
<!--     } -->
<!--   } -->

<!--   if (nrow(caaq_data) > 0) { -->
<!--     lineplot <- lineplot +  -->
<!--       geom_segment(data = caaq_data,  -->
<!--                    mapping = aes(x = as.Date(paste0(caaq_year_min, "-01-01")),  -->
<!--                                  xend = as.Date(paste0(caaq_year_max,"-12-31")),  -->
<!--                                  y = caaq_metric, yend = caaq_metric,  -->
<!--                                  colour = factor(caaq_status,  -->
<!--                                                  levels = c("Achieved", "Not Achieved"))),  -->
<!--                    size = 1.5) +  -->
<!--       annotate("text", x = as.Date(paste0(caaq_data$caaq_year_min, "-01-30")),  -->
<!--                y = 73, label = "2011-2013 Ozone Metric", size = 3.5, hjust=0,  -->
<!--                colour = "grey50") +  -->
<!--       geom_segment(data = caaq_data, colour = "grey60", -->
<!--                    aes(x = as.Date(paste0(caaq_year_min,"-09-15")), y = 69,  -->
<!--                        xend = as.Date(paste0(caaq_year_min, "-11-01")),  -->
<!--                        yend = caaq_metric + 1)) + -->
<!--       scale_colour_manual(values = c("#377eb8", "#e41a1c"), labels = "2011-2013 Ozone Metric",  -->
<!--                           name = element_blank(), guide = "none") -->
<!--   } -->

<!--   plot(lineplot) -->
<!--   cat("\\newpage ") -->
<!--   } -->
<!-- } -->

<!-- ``` -->


