---
topic: "air"
title: Status of Ground-Level Ozone in B.C. (2016-2018)
output: envreportutils.internal::print_ver
---
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, 
                      error=FALSE)

## Load libraries
library(knitr) #for knitting to PDF
library(english)
library(ggplot2) # for plotting
library(envreportutils) # to_titlecase()
library(purrr) #iwalk()
library(dplyr) #data munging

## Reading in plot objects
load("../tmp/plots.RData")

## Reading in results data objects for Appendix metadata
load("../tmp/analysed.RData")

## Set constants for 3-year reporting period
max_year <- 2018
min_year <- max_year - 2

## Label function for Appendix
knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # back slash has to be first
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}

```


British Columbia operates a network of air monitoring stations that
measure ground-level ozone and other air pollutants. This indicator
reports on the concentration of ground-level ozone from `r min_year`-`r max_year` and
compares it to the [Canadian Ambient Air Quality Standard](http://www.ccme.ca/en/resources/air/aqms.html#anchor1)
established by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

-   **Ground-level ozone is an air pollutant.** While ozone in the
    Earth's atmosphere occurs naturally, additional ozone at the
    ground level is a pollutant. It forms through chemical reactions
    involving nitrogen oxides and volatile organic compounds in the
    presence of sunlight.
-   **Ground-level ozone can be harmful to humans.** Exposure to
    ground-level ozone (hereafter ozone) can reduce lung function and
    cause inflammation of airways, which can increase respiratory
    symptoms and aggravate asthma^[1](#fn1)^. These effects are linked
    to more emergency room visits, hospitalizations, and absenteeism, and
    higher health care costs^[2](#fn2)^.
-   **Ozone levels met the Canadian Ambient Air Quality Standard at `r num_stns_achieved` of `r num_stns` B.C. 
    ozone monitoring stations.** Ozone levels (see box below) from the `r num_stns` 
    stations ranged from `r lowest_value` to `r highest_value` parts per billion (ppb).
    Ozone levels were $\leq\!\!$ 50 ppb at `r num_values_50` reporting stations (`r perc_values_50`%). `r to_titlecase(as.character(english(num_values_63)))`
    stations reported ozone levels \>63 ppb.
-   **Ozone levels met the Canadian Ambient Air Quality Standard in `r       as.character(english(num_az_achieved))` of B.C.'s seven air zones.** The ozone level     for an air zone is the highest ozone level reported from monitoring stations        within the air zone. Currently, there are no [air monitoring          stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest Air Zone.
-   **Ozone levels are used to set management levels for each air
    zone.** Four management levels (green, yellow, orange, and red) are each
    associated with a suite of actions that become more rigorous as
    ozone levels approach the Canadian Ambient Air Quality
    Standard.


\begin{center}
\fbox{
  \begin{minipage}[c]{15cm}
  
    \begingroup\centering{\textbf{Ground-Level Ozone: Canadian Ambient Air Quality Standard}}
    
    \endgroup
    
    \vspace{1mm}
    
    \text{Ozone levels are calculated using a statistical form called the ozone metric.}
    
    \text{The Canadian Ambient Air Quality Standard value for ozone is \textbf{63 parts per billion (ppb)}.} 
    
    \text{The ozone standard is achieved when ozone levels are $\leq$63 ppb.}
    
    
  \end{minipage}
}
\end{center}


\begin{center}
\fbox{
  \begin{minipage}[c]{15cm}
  
    \begingroup\centering{\textbf{What is an Air Zone?}}
    
    \endgroup
    
    \vspace{1mm}
    
    \text{British Columbia is divided up into seven air zones, which are areas that typically}
    
    \text{exhibit similar air quality characteristics, issues, and trends.}
    
  \end{minipage}
}
\end{center}



\newpage

The map and chart below summarise Canadian Ambient Air Quality Standard
achievement status for ozone in B.C. air zones, as well as ozone levels at individual monitoring stations. Summaries are given for each monitoring station where sufficient data was available for the `r min_year`-`r max_year`
reporting period.

# Status of Ground-Level Ozone Levels in B.C. Air Zones (`r min_year`-`r max_year`)

```{r achievement_map, fig.height=8, fig.width=7}

achievement_map

```

\newpage

# Ozone Metrics for Air Monitoring Stations within B.C. Air Zones (`r min_year`-`r max_year`)

```{r ambient_summary_plot, fig.height=9, fig.width=7}

ambient_summary_plot

```



### More about the Canadian Ambient Air Quality Standard ozone metric:

-   The statistical form of the Canadian Ambient Air Quality Standard ozone metric is the 3-year average of the annual 4th-highest daily maximum 8-hour rolling average concentration.
-   The ozone metric for an air zone is the highest ozone metric reported from monitoring stations within the air zone. Air zone metrics are only taken from stations reporting on two years of data when there are no available three year reporting stations in the air zone.
-   Only air monitoring stations with sufficient data on ozone
    concentrations for the `r min_year`-`r max_year` reporting period were included in
    this indicator. Visit [Current Air Quality Data](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E) for a
    complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the ozone
    metric value are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).

## Canada-wide Air Quality Management System

The Air Quality Management System is Canada's approach for protecting air quality. Under the Air Quality Management System, provinces and territories monitor, report and manage local air quality within air zones, with the goal of continuously improving air quality and keeping pollutants below the Canadian Ambient Air Quality Standards.

-   There are seven air zones in British Columbia: Coastal, Georgia Strait, Lower Fraser Valley, Southern Interior, Central Interior, Northeast and Northwest.
-   The Air Zone Management Framework defines management levels based on the highest ozone level reported from monitoring stations within the air zone. The four management levels are (1) green (actions for keeping clean areas clean); (2) yellow (actions for preventing air quality deterioration); (3) orange (actions for preventing Canadian Ambient Air Quality Standard exceedance); and (4) red (actions for achieving air zone Canadian Ambient Air Quality Standard). Actions for each management level are detailed in the [Guidance Document on Air Zone Management (2019) (PDF)](https://www.ccme.ca/files/Resources/air/Guidance%20Document%20on%20Air%20Zone%20Management.pdf). Air zone management levels are preferentially based on stations with 3 complete years of data.
-   The Coastal and Northeast air zones were assigned to a green management level, the Central Interior, Southern Interior and Georgia Strait air zones were assigned to a yellow management level, and the Lower Fraser Valley was assigned to an orange management level. Currently, there are no [air monitoring stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E) in the Northwest air zone.


# Management Levels for Ground-Level Ozone in B.C. Air Zones (`r min_year`-`r max_year`)

```{r management_map, fig.height = 5} 

management_map_pdf

```

\vspace{12pt}

```{r management_chart, fig.height=4}

management_chart +  theme(plot.margin = unit(c(10,30,0,0),"mm"),
                          axis.text = element_text(size = 10),
                          axis.title = element_text(size = 12),
                          legend.title = element_text(size = 12),
                          legend.text = element_text(size = 10))

```


### More about the Air Quality Management System management levels:

-   There can be cases when an air zone does not achieve a given Canadian Ambient Air Quality Standard
    because of sources over which jurisdictions have little or no
    control, such as those related to transboundary flows and
    exceptional events, like forest fires.
-   Prior to determining management levels, jurisdictions have the
    option of adjusting their air zone metric values to remove such
    external influences. These arrangements aim to ensure that
    jurisdictions are responsible for managing only the emissions
    sources they can control.
-   Wildfire influences were determined for two monitoring sites ([Hope](http://www.env.gov.bc.ca/epd/bcairquality/readings/map/station.html#E223756) and [Mission](http://www.env.gov.bc.ca/epd/bcairquality/readings/map/station.html#E302130)) in B.C. on a total of thirteen days for the `r min_year`-`r max_year` reporting period. The daily concentrations for these days were removed from the calculations of management levels, resulting in a change from red to orange for the Lower Fraser Valley management level for ground-level ozone.


# Methods

The methods used to develop this indicator---including procedures, data
requirements, and calculation of the Canadian Ambient Air Quality Standard ozone metric---are detailed in
the [Guidance Document on
Achievement Determination: Canadian Ambient Air Quality Standards for
Fine Particulate Matter and Ozone
(2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf) published by the [Canadian Council of Ministers of the Environment](http://www.ccme.ca/index.html).

**R package and code:** We have developed an [R](http://www.r-project.org) package to facilitate the calculation of air quality metrics according to the [Canadian Ambient Air Quality Standards](http://www.ccme.ca/en/current_priorities/air/caaqs.html). Download the 'rcaaqs' package from [GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating the analysis presented on this page is also available on [GitHub](https://github.com/bcgov/ozone_caaqs_analysis).


# References and Other Useful Links

Read individual [Air Zone reports](http://www2.gov.bc.ca/gov/search?id=04D14E7F7EE043889E1AC20E205CA3F8&tab=1&sourceId=267EE7E2432B4DE89BF1905EF80F505E&q=%22air+zone+report%22) on the achievement of the Canadian Ambient Air Quality Standards for ground-level ozone and fine particulate matter in B.C.

[Learn more about the implementation of the Air Quality Management System in British Columbia](http://www2.gov.bc.ca/gov/content?id=8FA84ADC845A49AB93E525A58D9D4A97)

[Access B.C.'s real-time data on air pollutants and find locations of all provincial air monitoring stations in B.C.](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)

[BC Lung Asssociation's BC State of the Air Reports](http://www.bc.lung.ca/airquality/stateoftheair-report.html)

[Canadian Environmental Sustainability Indicators: Air and Climate Indicators](http://www.ec.gc.ca/indicateurs-indicators/default.asp?lang=En&n=03603FB3-1)

^1^[United States Environmental Protection Agency. February 2013.
Integrated Science Assessment of Ozone and Related Photochemical
Oxidants
(600/R-10/076F)](http://cfpub.epa.gov/ncea/isa/recordisplay.cfm?deid=247492#Download)

^2^Willey, J., N. Gilbert, and N. Lyrette. 2004. Human health effects of
ozone: Update in support of the Canada-wide standards for particulate
matter and ozone. Revised version. Working paper prepared for Canadian
Council of Ministers of the Environment. Health Canada. Ottawa, Ontario:
Health Canada

# Data

\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Indicator Summary Data: BC Ground-Level Ozone Canadian Ambient Air Quality Standards `r min_year`-`r max_year`](https://catalogue.data.gov.bc.ca/dataset/9b7a9e74-9274-4f97-be81-ce4ee475077d) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [B.C. Ground-level Ozone Hourly Data & Air Monitoring Station Locations](https://catalogue.data.gov.bc.ca/dataset/77eeadf4-0c19-48bf-a47a-fa9eef01f409) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [B.C. Air Zones](https://catalogue.data.gov.bc.ca/dataset/e8eeefc4-2826-47bc-8430-85703d328516) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (April 2021):
<http://www.env.gov.bc.ca/soe/indicators/air/ozone.html>

Email correspondence to: envreportbc@gov.bc.ca

*Suggested Citation*:  
Environmental Reporting BC. 2021. Status of Ground-Level Ozone in B.C. (`r min_year`-`r max_year`). State of Environment Reporting, Ministry of Environment and Climate Change Strategy, British Columbia, Canada.


\newpage

# Appendix: Detailed Ground-Level Ozone Canadian Ambient Air Quality Standard Results for Monitoring Stations within Air Zones in B.C. (`r min_year`-`r max_year`)


```{r appendix, results='asis', fig.height=3.5}

iwalk(stn_plots,  ~ {
  
  stn <- ozone_caaqs_results %>% 
    filter(ems_id == .y) 
  
  az <- stn %>% pull(airzone)
  cat("\\textbf{Air Zone:} ", az, " \\newline ")
  
  stn_name <- stn %>% pull(station_name)
  site <- knitr_latex_char(stn_name)
  cat("\\textbf{Monitoring Station:} ", site, " \\newline ")
  
  status <- stn %>% pull(caaqs_ambient) %>% as.character()
  cat("\\textbf{Ozone Air Quality Standard:} ",
      status, " \\newline ")
  
  value <- stn %>% pull(metric_value_ambient)
  nys <- stn %>% pull(n_years)
  cat("\\textbf{Ozone Metric:} ", value, " ppb (", nys,
      " year average) \\newline ", " \\newline ", sep = "")
  
  plot <- .x
  plot(plot)
  cat("\\newpage ")
  
})

```


