---
title: Air
subtitle: Status of Ground-Level Ozone in B.C. (2011-2013)
output:
  pdf_document:
    template: D:/templates/print_ver_template.tex
---

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, 
                      error=FALSE, fig.width=7)

library("dplyr")
library("sp")
library("rgeos")
library("ggplot2")
library("grid")
library("scales")
library("envreportutils")

load("../tmp/analysed.RData")

ozone_sites <- ozone_sites[!is.na(ozone_sites$caaq_metric),]

min_year <- max_year - 2
maxdate <- as.Date(paste0(max_year, "-12-31"))
mindate <- as.Date(paste0(min_year, "-01-01"))

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesdn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}
```


British Columbia operates a network of air monitoring stations that
measure ground-level ozone and other air pollutants. This indicator
reports on the concentration of ground-level ozone from 2011-2013 and
compares it to the [Canadian Ambient Air Quality Standards
(CAAQS)](http://www.ccme.ca/en/current_priorities/air/caaqs.html)
developed by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

\begin{center}
\fbox{
  \begin{minipage}[c]{12cm}
    \centering{\bf Ground-Level Ozone\\ Canadian Ambient Air Quality Standard (CAAQS)}
    \vspace{-1mm}
    \begin{itemize}
      \itemsep1pt\parskip0pt\parsep0pt
     \item
     The CAAQS ozone metric should be \textbf{lower than 63 parts per billion}. 
     \item
     The \textbf{CAAQS ozone metric} is the 3-year average of the annual 4th-highest
     daily maximum 8-hour rolling average concentration of ozone.
      \end{itemize}
      \href{www.ccme.ca/en/current_priorities/air/caaqs.html}{Learn more about CAAQS ozone metric calculations.}
  \end{minipage}
}
\end{center}


-   **Ground-level ozone is an air pollutant.** While ozone in the
    Earth's atmosphere occurs naturally, additional ozone at the
    ground level is a pollutant. It forms through chemical reactions
    involving nitrogen oxides and volatile organic compounds in the
    presence of sunlight.
-   **Ground-level ozone can be harmful to humans.** Exposure to
    ground-level ozone (hereafter ozone) can reduce lung function and
    cause inflammation of airways, which can increase respiratory
    symptoms and aggravate asthma^[1](#fn1)^. These effects are linked
    to more emergency room visits, hospitalizations, and absenteeism, and
    higher health care costs^[2](#fn2)^.
-   **Ozone levels met the Canadian Ambient Air Quality Standard at all B.C. 
    ozone monitoring stations.** The CAAQS ozone metric
    from the 37 stations ranged from 38 to 60 parts per billion (ppb).
    Metric values were $\leq\!\!$ 50 ppb at 25 reporting stations (68%). No
    stations reported CAAQs metric values \>60 ppb.
-   **Ozone levels met the Canadian Ambient Air Quality Standard in four
    of B.C.'s seven air zones.** There was not enough data to estimate
    the CAAQS ozone metric for the remaining three air zones for this
    reporting period. The ozone level for an air zone
    is the highest CAAQS ozone metric value reported from monitoring
    stations within the air zone. 
-   **CAAQS ozone metrics are used to set management levels for each air
    zone.** Four management levels (green, yellow, orange, and red) are each
    associated with a suite of actions that become more rigorous as
    ozone metric values approach the Canadian Ambient Air Quality
    Standard.


The map and chart below summarise the Canadian Ambient Air Quality Standard (CAAQS)
achievement status for ozone in B.C. air zones, as well as the individual 
monitoring stations across the province. Summaries are given for each
monitoring station where sufficient data was available for the 2011-2013
reporting period. 

```{r achievement_map, fig.height=8}
airzones.points <- fortify(airzone_map, region = "Airzone") %>%
  rename(Airzone = id) %>%
  left_join(airzone_map@data, by = "Airzone")

stations.points <- as.data.frame(ozone_sites)

achievement_map <- ggplot(airzones.points, aes(long, lat)) + 
  geom_polygon(aes(group=group, fill = caaq_status)) + 
  coord_fixed() + 
  scale_fill_manual(values = c("Achieved" = "#377eb8", "Not Achieved" = "#e41a1c", 
                               "Insufficient Data" ="grey80"), drop = FALSE, 
                    name = "Airzones:\nOzone Air Quality Standard", 
                    guide = guide_legend(order = 1, title.position = "top")) + 
  geom_path(aes(group=group), colour = "white") + 
  geom_point(data = stations.points, aes(x = longitude, y = latitude, 
                                         colour = caaq_metric)) +
  scale_colour_gradient(high = "#252525", low = "#f0f0f0", 
                        name = "Monitoring Stations:\nOzone Metric (ppb)", 
                        guide = guide_colourbar(order = 2, title.position = "top", 
                                                barwidth = 10)) + 
  ggtitle("Status of Ozone Levels in B.C. Air Zones (2011-2013)") + 
  theme_minimal() + 
  theme(axis.title = element_blank(), axis.text = element_blank(), 
        axis.ticks = element_blank(), panel.grid = element_blank(), 
        legend.position = "bottom", legend.box = "horizontal", 
        legend.box.just = "top", legend.direction = "horizontal", 
        legend.title.align = 0, legend.margin = unit(20, "mm"),
        legend.title = element_text(face="plain", size = 11))

plot(achievement_map)
```

```{r summary-plot, fig.height=7}
sum_dat <- ozone_sites@data
sum_dat$Airzone <- reorder(sum_dat$Airzone, sum_dat$caaq_metric, max, order = TRUE)
sum_dat$Airzone <- factor(sum_dat$Airzone, levels = rev(levels(sum_dat$Airzone)))

ann_text <- data.frame(display_name = "Elk Falls Dogwood", caaq_metric = 62.5, 
                       Airzone = factor("Georgia Strait", 
                                        levels = levels(sum_dat$Airzone)))

summary_plot <- ggplot(sum_dat, 
                       aes(x = caaq_metric, 
                           y = reorder(display_name, caaq_metric, sum))) + 
  facet_grid(Airzone~., scales = "free_y", space = "free_y", drop=TRUE, 
             labeller = label_wrap_gen(15)) + 
  geom_point(size = 2, colour = "#377eb8") + 
  geom_vline(aes(xintercept = o3_standard), linetype = 2, colour = "#e41a1c") + 
  labs(title="Ozone Metric for Air Monitoring Stations within B.C. Air Zones",
       x = "CAAQS Ozone Metric", y = "Ozone Monitoring Station") + 
  geom_text(data = ann_text, label = "Ozone Standard (63 ppb)", size = 3, 
            hjust = 1, colour = "#e41a1c") + 
  theme_soe_facet(base_family = "", use_sizes=FALSE) + 
  theme(axis.title = element_text(size=rel(1)), 
        axis.text.y = element_text(size = rel(0.7)), 
        axis.line.x = element_blank(), 
        strip.text = element_text(size = rel(0.75)))

plot(summary_plot)

```


### More about the CAAQS ozone metric: 

-   The CAAQS ozone metric is measured as the 3-year average of the
    annual 4th-highest daily maximum 8-hour rolling average
    concentration.
-   The ozone level for an air zone is the highest CAAQS ozone metric value
    reported from monitoring stations within the air zone. However, stations 
    with metric values based on only two years of data are excluded for
    consideration of the *air zone* metric value.
-   Only air monitoring stations with sufficient data on ozone
    concentrations for the 2011-2013 reporting period were included in
    this indicator. Visit [B.C Air
    Quality](http://www.bcairquality.ca/readings/index.html) for a
    complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the CAAQS ozone
    metric are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).

## Canada-wide Air Quality Management System

The Air Quality Management System (AQMS) is Canada's approach for
protecting air quality. Under the AQMS, provinces and territories
monitor, report and manage local air quality within air
zones, with the goal of continuously improving air quality and
keeping pollutants below the Canadian Ambient Air Quality Standards
(CAAQS).

-   Provinces and territories establish air zones that are defined by a
    similar set of air quality characteristics, issues and trends. 
-   There are seven air zones in British Columbia: Coastal, Georgia Strait,
    Lower Fraser Valley, Southern Interior, Central Interior, Northeast
    and Northwest.
-   The Air Zone Management Framework defines management levels based on
    the highest CAAQS ozone metric value reported from monitoring stations
    within the air zone.
-   The four management levels include: (1) green (actions for keeping
    clean areas clean); (2) yellow (actions for preventing air quality
    deterioration); (3) orange (actions for preventing CAAQS
    exceedance); and (4) red (actions for achieving air zone CAAQS). Each management level is     associated with a suite of actions that become more rigorous as ozone concentration levels     approach
    the CAAQS.
-   All four B.C. air zones that were assessed met the Canadian Ambient Air
    Quality Standard for ground-level ozone ($\leq\!\!$ 63 ppb).
-   Based on the highest CAAQS ozone metric value reported within each air zone, 
    the Central Interior, Georgia Strait, and Lower Fraser Valley air zones 
    were assigned to a yellow management level (actions for preventing air 
    quality deterioration) and the Southern Interior air zone to an orange 
    management level (actions for preventing CAAQS exceedance). Due to 
    insufficient data, ozone management levels in the Coastal, Northwest and 
    Northeast air zones are yet to be determined.
-   Actions for each management level are detailed in the
    [CCME](http://www.ccme.ca/en/current_priorities/air/caaqs.html)
    [Guidance Document on Air Zone Management
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf).
   
```{r mgmt_map, fig.height = 5}
colrs <- c("Insufficient Data" = "grey80", 
           "Actions for Keeping Clean Areas Clean" = "#A6D96A", 
           "Actions for Preventing Air Quality Deterioration" = "#FEE08B", 
           "Actions for Preventing CAAQS Exceedance" = "#F46D43", 
           "Actions for Achieving Air Zone CAAQS" = "#A50026")

mgmt_map <- ggplot(airzones.points, aes(long, lat)) +   
  geom_polygon(aes(group=group, fill = caaq_level)) + 
  coord_fixed() + 
  geom_path(aes(group=group), colour = "white") + 
  ggtitle("Management Levels for Ozone in B.C. Air Zones") + 
  theme_minimal() + 
  scale_fill_manual(values = colrs, 
                    drop = FALSE, 
                    name = "Air Zone Management Levels", 
                    guide = guide_legend(reverse = TRUE)) + 
  theme(axis.title = element_blank(),
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
       legend.position = "none",
       plot.margin = unit(c(0,0,0,0),"mm")) +
  annotate("text", x=680000, y=950000, label="Coastal", 
           colour="black", size=3.5) +
  annotate("text", x=1150000, y=1550000, label="Northeast", 
           colour="black", size=3.5) +
  annotate("text", x=780000, y=1500000, label="Northwest", 
           colour="black", size=3.5) +
  annotate("text", x=1150000, y=950000, label="Central\n Interior", 
           colour="black", size=3.5) +
  annotate("text", x=1550000, y=600000, label="Southern\n Interior", 
           colour="black", size=3.5) +
  annotate("text", x=1150000, y=325000, label="Georgia Strait", 
           colour="black", size=3.5) +
  annotate("text", x=1500000, y=410000, label="Lower Fraser Valley", 
           colour="black", size=3.5)

plot(mgmt_map)
```

```{r mgmt_chart, fig.height=4}
chart_colrs <- colrs[-1]

# <- c(expression(paste("Actions for Keeping Clean Areas Clean ("<=50, " ppb)")), 
#             expression(paste("Actions for Preventing Air Quality Deterioration (">50, " & " <= 56, " #ppb)")), 
#             expression(paste("Actions for Preventing CAAQS Exceedance (">56, " & " <=63, " ppb)")), 
#             expression(paste("Actions for Achieving Air Zone CAAQS (">63, " ppb)")))

mlevels <- c("Actions for Keeping Clean Areas Clean (<=50 ppb)", 
             "Actions for Preventing Air Quality Deterioration (>50 & <=56 ppb)", 
             "Actions for Preventing CAAQS Exceedance (>56 & <=63 ppb)", 
             "Actions for Achieving Air Zone CAAQS (>63 ppb)")

mgmt_chart <- ggplot(data=stations.points,
                              aes(x = Airzone, fill = caaq_level)) + 
  geom_bar(stat = "bin", alpha = 1) +
  xlab ("") + ylab ("Number of Reporting Stations") +
  scale_y_continuous(limits = c(0,20), breaks=seq(0, 20, 4),
                     expand=c(0,0)) +
  scale_fill_manual(values = chart_colrs, 
                    drop = FALSE, 
                    name = "Air Zone Management Levels", 
                    guide = guide_legend(reverse = TRUE),
                    labels = mlevels) + 
  theme_soe(base_family = "") +
  theme(panel.grid.major.y = (element_blank()),
        axis.text = element_text(size = 12),
        legend.position = "bottom",
        legend.direction = "vertical",
        legend.margin = unit(15,"mm"),
        legend.text.align = 0) +
  coord_flip()

plot(mgmt_chart)
```


### More about the AQMS management levels:

-   There can be cases when an air zone does not achieve a given CAAQS
    because of sources over which jurisdictions have little or no
    control, such as those related to transboundary flows and
    exceptional events like forest fires.
-   Prior to determining management levels, jurisdictions have the
    option of adjusting their air zone metric values to remove such
    external influences. These arrangements aim to ensure that
    jurisdictions are responsible for managing only the emissions
    sources they can control.
-   For the 2011-2013 reporting period, all ozone monitoring stations met
    the CAAQS and no adjustments for transboundary or exceptional events were
    warranted. Air zone management levels for ground-level ozone were
    determined directly from the CAAQS ground-level ozone metric for this
    reporting period. 

# Methods

The methods used to develop this indicator --- including procedures, data
requirements, and calculation of the CAAQS ozone metric --- are detailed in
the [CCME](http://www.ccme.ca/index.html) [Guidance Document on
Achievement Determination: Canadian Ambient Air Quality Standards for
Fine Particulate Matter and Ozone
(2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).

# References and Other Useful Links

Visit [BC Air Quality](http://www.bcairquality.ca/plans/national-air-quality-management-system.html) for more information on the implementation of the AQMS in B.C., and to read individual Air Zone reports on the achievement of CAAQS for ground-level ozone and fine particulate matter.

Access [B.C.'s Air Quality
Readings](http://www.bcairquality.ca/readings/index.html) for real-time
data on air pollutants and locations of all air monitoring stations in
B.C.

For more details on the CAAQS and Canada's Air Quality Management System
visit [Canada-wide Air Quality Management System
(AQMS)](http://www.ccme.ca/en/resources/air/aqms.html)

[BC Lung Asssociation's BC State of the Air Reports](http://www.bc.lung.ca/airquality/stateoftheair-report.html)

^1^[United States Environmental Protection Agency. February 2013.
Integrated Science Assessment of Ozone and Related Photochemical
Oxidants
(600/R-10/076F)](http://cfpub.epa.gov/ncea/isa/recordisplay.cfm?deid=247492#Download)

^2^Willey, J., N. Gilbert, and N. Lyrette. 2004. Human health effects of
ozone: Update in support of the Canada-wide standards for particulate
matter and ozone. Revised version. Working paper prepared for Canadian
Council of Ministers of the Environment. Health Canada. Ottawa, Ontario:
Health Canada

# Data

\*By accessing these datasets, you agree to the license associated with
each file, as indicated in parentheses below.

-   [Indicator data: BC Ozone CAAQS 2011-2013](http://catalogue.data.gov.bc.ca/dataset/bc-ambient-air-quality-standards-reporting-results-ozone) ([License: B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [Hourly ozone data](http://catalogue.data.gov.bc.ca/dataset/air-quality-monitoring-verified-hourly-data-and-station-data) (License: [B.C.     OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [BC Air Monitoring     Stations](http://catalogue.data.gov.bc.ca/dataset/air-quality-monitoring-verified-hourly-data-and-station-data)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [BC Air Zones](http://catalogue.data.gov.bc.ca/dataset/british-columbia-air-zones) (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))

Published and Available On-Line at Environmental Reporting BC (August 2015):
<http://www.env.gov.bc.ca/soe/indicators/air/ozone.html>

Email correspondence to: envreportbc@gov.bc.ca

\newpage

# Appendix: Detailed CAAQS results for each monitoring station within air zones in B.C.

```{r appendix, results='asis', fig.height=3.5}
plot_exceedances <- FALSE

for (az in (unique(sum_dat$Airzone))) {
  dat <- sum_dat[sum_dat$Airzone == az, ]
  ems_ids <- unique(dat$ems_id)
  cat("\\subsection{", az, " Air Zone}", sep = "")

  for (emsid in ems_ids) {
    site_sum <- dat[dat$ems_id == emsid, ]
    dailydata <- daily_8hr_roll_max[daily_8hr_roll_max$ems_id == emsid &
                                      daily_8hr_roll_max$date <= maxdate,]

    if (nrow(dailydata) == 0) next

    site <- knitr_latex_char(site_sum$display_name[1])
    cat("\\subsubsection*{\\underline{", site, " monitoring station}}", sep = "")
    cat("\\textbf{Ozone Air Quality Standard:} ", site_sum$caaq_status, " \\newline ")
    cat("\\textbf{Ozone Metric:} ", site_sum$caaq_metric, " ppb (", site_sum$caaq_nYears, 
        " year average) \\newline ", sep = "")
    cat("\\textbf{Air Quality Management Level:} ", as.character(site_sum$caaq_level), 
        " \\newline \\newline \\newline ")

    caaq_data <- three_yr_avg[three_yr_avg$ems_id == emsid,]

  lineplot <- ggplot(dailydata, size = 1) + 
    scale_x_date(expand = c(0,50), limits = c(mindate - 1, maxdate), 
                 breaks = date_breaks(width = "1 year"), labels = date_format("%Y")) + 
  scale_y_continuous(limits = c(0, 102), 
                     breaks = seq(0, 100, by = 20), labels = seq(0, 100, by = 20), 
                     expand = c(0,0)) + 
  geom_line(aes(x = date, y = max8hr), colour = "#9ecae1", size = 0.5) + 
    geom_hline(aes(yintercept = o3_standard), linetype = 2, colour = "#e41a1c") + 
    annotate("text", label = paste0("Ozone Standard (", o3_standard, " ppb)  "), 
             x = maxdate, y = o3_standard + 3, vjust = 0, hjust = 1, 
             size = 3.5, colour = "#e41a1c") + 
  theme_soe(base_family = "", base_size = 10) + 
    theme(axis.title.y = element_text(vjust = 1)) + 
    labs(x = NULL, y = "Daily Maximum Ozone\n(parts per billion)")
  
  if (plot_exceedances) {
    exceedance_data <- filter(dailydata, exceed)
    
    if (nrow(exceedance_data) > 0) {
      lineplot <- lineplot + 
        geom_point(data = exceedance_data, aes(x = date, y = max8hr), 
                   colour = "#e41a1c", size = 2) + 
        annotate("text", x = exceedance_data$date[1] + 20, y = exceedance_data$max8hr[1], 
                 label = "Exceedances", hjust = 0, vjust = 0, colour = "#e41a1c", size = 3)
    }
  }
  
  if (nrow(caaq_data) > 0) {
    lineplot <- lineplot + 
      geom_segment(data = caaq_data, 
                   mapping = aes(x = as.Date(paste0(caaq_year_min, "-01-01")), 
                                 xend = as.Date(paste0(caaq_year_max,"-12-31")), 
                                 y = caaq_metric, yend = caaq_metric, 
                                 colour = factor(caaq_status, 
                                                 levels = c("Achieved", "Not Achieved"))), 
                   size = 1.5) + 
      annotate("text", x = as.Date(paste0(caaq_data$caaq_year_min, "-01-30")), 
               y = 73, label = "2011-2013 Ozone Metric", size = 3.5, hjust=0, 
               colour = "grey50") + 
      geom_segment(data = caaq_data, colour = "grey60",
                   aes(x = as.Date(paste0(caaq_year_min,"-09-15")), y = 69, 
                       xend = as.Date(paste0(caaq_year_min, "-11-01")), 
                       yend = caaq_metric + 1)) +
      scale_colour_manual(values = c("#377eb8", "#e41a1c"), labels = "2011-2013 Ozone Metric", 
                          name = element_blank(), guide = "none")
  }
  
  plot(lineplot)
  cat("\\newpage ")
  }
}

```


